---
title: "Amplicon sequence data pre-processing"
author: "Srotoswini Joardar"
date: "`r Sys.Date()`"
output:
  github_document:
  toc_depth: 3
  number_sections: false
  toc: true
---

# Script overview

This script performs R-based downstream processing and quality filtering of ASV data following Galaxy-based preprocessing. Raw demultiplexed FASTQ files were processed using AdapterRemoval and DADA2 for quality filtering, read merging, chimera removal, ASV inference, and taxonomic assignment against SILVA v138.2. Resulting ASV and taxonomy tables from all sequencing runs were imported for downstream analysis. 

Quality filtering thresholds were defined based on read quality profiles, with forward reads trimmed at position 200 (maxEE = 10) and reverse reads at position 270 (maxEE = 8).

Raw sequencing data are available in the NCBI Sequence Read Archive (SRA) BioProject: PRJNA1406503.
This script begins with the post-Galaxy outputs such as ASV tables (post trimming and filtering) and taxonimic assignment table.


# 0. Setting up and loading packages
```{r, message = FALSE, warning = FALSE}

set.seed(42)

library(readxl)
library(dplyr)
library(microbiome)
library(phyloseq)
library(ggplot2)
library(scales)
library(decontam)
library(tibble)
library(dada2)
library(vegan)
library(metagMisc)
```

# 1. Importing metadata

To initiate the workflow, a metadata file containing all relevant information about the biological samples must first be created in Excel format
```{r}

metadata <- read_xlsx("metadata_wini.xlsx")
metadata <- as.data.frame(metadata)
```

# 2. Loading counts table and changing row names

In this section, the read count tables for all the runs generated by the Galaxy workflow is imported for downstream analysis
```{r}

Pool1_counts <- read.csv("makeSequenceTable pool1.csv")
Pool1_counts <- as.data.frame(Pool1_counts)
Pool1_counts <- Pool1_counts %>%
  column_to_rownames("ASV")

Pool2_counts <- read.csv("makeSequenceTable pool2.csv")
Pool2_counts <- as.data.frame(Pool2_counts)
Pool2_counts <- Pool2_counts %>%
  column_to_rownames("ASV")

Pool3_counts <- read.csv("makeSequenceTable pool3.csv")
Pool3_counts <- as.data.frame(Pool3_counts)
Pool3_counts <- Pool3_counts %>%
  column_to_rownames("ASV")

posC_counts <- read.csv("makeSequenceTable posC.csv")
posC_counts <- as.data.frame(posC_counts)
posC_counts <- posC_counts %>%
  column_to_rownames("ASV")
```

# 3. Adding 0 to any empty field
```{r}

Pool1_counts[is.na(Pool1_counts)] <- 0

Pool2_counts[is.na(Pool2_counts)] <- 0

Pool3_counts[is.na(Pool3_counts)] <- 0

posC_counts[is.na(posC_counts)] <- 0
```

# 4. Converting to matrix
```{r}

Pool1_counts <- Pool1_counts %>%
  as.matrix() %>% t()

Pool2_counts <- Pool2_counts %>%
  as.matrix()  %>% t()

Pool3_counts <- Pool3_counts %>%
  as.matrix() %>% t()

posC_counts <- posC_counts %>%
  as.matrix()  %>% t()
```

# 5. Creating combined OTU table
```{r}

Big_OTU <- dada2::mergeSequenceTables(Pool1_counts, 
                                      Pool2_counts, 
                                      Pool3_counts, 
                                      posC_counts, 
                                      repeats = "sum", 
                                      orderBy = NULL) %>% t()
```

# 6. Loading taxonomy tables

The taxonomy assignment tables generated by the Galaxy workflow are imported in this section
```{r}

Pool1_tax <- read.csv("assignTaxonomy pool1.csv")
Pool2_tax <- read.csv("assignTaxonomy pool2.csv")
Pool3_tax <- read.csv("assignTaxonomy pool3.csv")
posC_tax <- read.csv("assignTaxonomy posC.csv")
```

# 7. Merging taxonomy tables
```{r, message = FALSE, warning = FALSE}

Tax_all <- rbind(Pool1_tax, 
                 Pool2_tax, 
                 Pool3_tax, 
                 posC_tax) %>% distinct()

tax_name <- c("asv","Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")

colnames(Tax_all) <- tax_name
```

# 8. Removing duplicate rows
```{r}

##Extracting duplicate rows with conflicting taxonomic information
duplicate_rows <- Tax_all %>%
  group_by(asv) %>%
  filter(n() > 1)

##Prioritizing rows with fewest NAs
most_info_rows <- duplicate_rows %>%
  rowwise() %>%
  mutate(missing_count = sum(is.na(c_across(all_of(colnames(Tax_all[-1])))))) %>%
  group_by(asv) %>%
  filter(missing_count == min(missing_count)) %>%
  ungroup() %>%
  select(-missing_count)

##Kicking out the duplicated rows and adding rows with most information
bound_less <- Tax_all %>%
  filter(!(asv %in% duplicate_rows$asv))

Tax_all <- rbind(bound_less, most_info_rows) %>%
  column_to_rownames("asv") %>%
  as.matrix() 
```

# 9. Creating phyloseq object
```{r}

rownames(metadata) <- metadata[,1]
sample_data <- metadata[,-1]

OTU = otu_table(Big_OTU, taxa_are_rows = TRUE)
TAX = tax_table(Tax_all)
samples = sample_data(metadata)

Phylo_Obj <- phyloseq(OTU, TAX, samples)
```

# 10. Removing singletons
```{r}

Phylo_Obj<-prune_taxa(taxa_sums(Phylo_Obj)> 1, Phylo_Obj)
```

# 11. Splitting different compartments to remove contaminants
```{r}

Roots_ps <- subset_samples(Phylo_Obj, Compartment == "Root")  
sum(Roots_ps@otu_table)

R_T0_ps <- subset_samples(Roots_ps, Time_point == "T0")  
sum(R_T0_ps@otu_table) 

R_T1_ps <- subset_samples(Roots_ps, Time_point == "T1")  
sum(R_T1_ps@otu_table)

R_T2_ps <- subset_samples(Roots_ps, Time_point == "T2")  
sum(R_T2_ps@otu_table)
```

# 12. Remove NAs from table
```{r}

R_T0_ps2 <- transform_sample_counts(R_T0_ps, 
                                    function(x) ifelse(is.na(x), 0, x))
sum(R_T0_ps2@otu_table)

R_T1_ps2 <- transform_sample_counts(R_T1_ps, 
                                    function(x) ifelse(is.na(x), 0, x))
sum(R_T1_ps2@otu_table)

R_T2_ps2 <- transform_sample_counts(R_T2_ps, 
                                    function(x) ifelse(is.na(x), 0, x))
sum(R_T2_ps2@otu_table)
```

# 13. Prune samples where the sum is greater than 0 for each dataset
```{r}

R_T0_ps3 <- prune_samples(sample_sums(R_T0_ps2) > 0, R_T0_ps2)

R_T1_ps3 <- prune_samples(sample_sums(R_T1_ps2) > 0, R_T1_ps2)

R_T2_ps3 <- prune_samples(sample_sums(R_T2_ps2) > 0, R_T2_ps2)
```

# 14.1. Ordinate using PCoA with Bray-Curtis distance for each dataset
```{r}

Control_ord_t0 <- ordinate(R_T0_ps3, 
                           method = "PCoA", 
                           distance = "bray")

Control_ord_t1 <- ordinate(R_T1_ps3, 
                           method = "PCoA", 
                           distance = "bray")

Control_ord_t2 <- ordinate(R_T2_ps3, 
                           method = "PCoA", 
                           distance = "bray")
```

## 14.2.Plot ordination for each dataset

This step allows to check the number of samples and also the controls
```{r, message = FALSE, warning = FALSE}

p1_t0 <- plot_ordination(R_T0_ps3, Control_ord_t0, 
                         type = "Sample", 
                         color = "Type")
p1_t1 <- plot_ordination(R_T1_ps3, Control_ord_t1, 
                         type = "Sample", 
                         color = "Type")
p1_t2 <- plot_ordination(R_T2_ps3, Control_ord_t2, 
                         type = "Sample", 
                         color = "Type")

##Print the plots
p1_t0
p1_t1
p1_t2
```

# 15. Summing the OTU table for each dataset

This step is to check the number of reads before any of the removal steps. After each step of contaminant/control/chloroplast/mitochondria removal, the sum function was used to document the number of reads remaining after each round of removal
```{r}

sum(R_T0_ps3@otu_table) 
sum(R_T1_ps3@otu_table) 
sum(R_T2_ps3@otu_table) 
```

# 16. Identify contaminants with prevalence method
```{r}

sample_data(R_T0_ps3)$is.neg <- sample_data(R_T0_ps3)$Type == "Control"
contam_t0_df <- isContaminant(R_T0_ps3, 
                              method = "prevalence", 
                              neg = "is.neg", 
                              threshold = 0.5)
table(contam_t0_df$contaminant)

contam_t0_asvs <- row.names(contam_t0_df[contam_t0_df$contaminant == TRUE, ])

sample_data(R_T1_ps3)$is.neg <- sample_data(R_T1_ps3)$Type == "Control"
contam_t1_df <- isContaminant(R_T1_ps3, 
                              method = "prevalence", 
                              neg = "is.neg", 
                              threshold = 0.5)
table(contam_t1_df$contaminant)

contam_t1_asvs <- row.names(contam_t1_df[contam_t1_df$contaminant == TRUE, ])

sample_data(R_T2_ps3)$is.neg <- sample_data(R_T2_ps3)$Type == "Control"
contam_t2_df <- isContaminant(R_T2_ps3, 
                              method = "prevalence", 
                              neg = "is.neg",
                              threshold = 0.5)
table(contam_t2_df$contaminant)

contam_t2_asvs <- row.names(contam_t2_df[contam_t2_df$contaminant == TRUE, ])
```

# 17. Filtering contaminants
```{r}

ps_filt_t0 <- prune_taxa(!taxa_names(R_T0_ps3) %in% contam_t0_asvs, R_T0_ps3)
sum(ps_filt_t0@otu_table) 

ps_filt_t1 <- prune_taxa(!taxa_names(R_T1_ps3) %in% contam_t1_asvs, R_T1_ps3)
sum(ps_filt_t1@otu_table) 

ps_filt_t2 <- prune_taxa(!taxa_names(R_T2_ps3) %in% contam_t2_asvs, R_T2_ps3)
sum(ps_filt_t2@otu_table) 
```

# 18. Removing negative controls and positive controls from the dataset
```{r}

ps_filt_t0 <- subset_samples(ps_filt_t0, Type == "Sample")
sum(ps_filt_t0@otu_table) 

ps_filt_t1 <- subset_samples(ps_filt_t1, Type == "Sample")
sum(ps_filt_t1@otu_table) 

ps_filt_t2 <- subset_samples(ps_filt_t2, Type == "Sample")
sum(ps_filt_t2@otu_table)
```

# 19. Removing Chloroplasts and Mitochondria
```{r}

ps_filt_t0 <- subset_taxa(ps_filt_t0, Order != "Chloroplast")
sum(ps_filt_t0@otu_table) 
ps_filt_t0 <- subset_taxa(ps_filt_t0, Family != "Mitochondria")
sum(ps_filt_t0@otu_table) 

ps_filt_t1 <- subset_taxa(ps_filt_t1, Order != "Chloroplast")
sum(ps_filt_t1@otu_table) 
ps_filt_t1 <- subset_taxa(ps_filt_t1, Family != "Mitochondria")
sum(ps_filt_t1@otu_table) 

ps_filt_t2 <- subset_taxa(ps_filt_t2, Order != "Chloroplast")
sum(ps_filt_t2@otu_table) 
ps_filt_t2 <- subset_taxa(ps_filt_t2, Family != "Mitochondria")
sum(ps_filt_t2@otu_table)
```

# 20. Rarefaction curves
```{r}

t0_Otu <- as.data.frame(ps_filt_t0@otu_table)

raret0 <- rarecurve(t(t0_Otu[1:43]), 
                   step = 50,  
                   lwd = 2, 
                    xlim = c(0,80000), 
                    ylab = "ASVs", 
                    xlab = "Number of Reads", 
                    label = FALSE , 
                    cex = 0.5,
                    title(main = "Rarefaction Curve for Roots at T0", 
                          col.main = "black", 
                          font.main = 2))
 
t1_Otu <- as.data.frame(ps_filt_t1@otu_table)

raret1 <- rarecurve(t(t1_Otu[1:47]), 
                     step = 50,  
                     lwd = 2, 
                     xlim = c(0,80000), 
                     ylab = "ASVs", 
                     xlab = "Number of Reads", 
                     label = FALSE , 
                     cex = 0.5,
                     title(main = "Rarefaction Curve for Roots at T1", 
                           col.main = "black", 
                           font.main = 2))

t2_Otu <- as.data.frame(ps_filt_t2@otu_table)

raret2 <- rarecurve(t(t2_Otu[1:46]), 
                     step = 50,  
                     lwd = 2, 
                     xlim = c(0,80000), 
                     ylab = "ASVs", 
                     xlab = "Number of Reads", 
                     label = FALSE , 
                     cex = 0.5,
                     title(main = "Rarefaction Curve for Roots at T2", 
                           col.main = "black", 
                           font.main = 2))
```

# 21. Merging all phyloseq objects
```{r, eval = FALSE}

merged_physeq <- merge_phyloseq(ps_filt_t0, ps_filt_t1, ps_filt_t2)
tax_table(merged_physeq) <- cbind(tax_table(merged_physeq), 
                                  Sequence = rownames(tax_table(merged_physeq)))
merged_physeq <- add_refseq(merged_physeq, tag = "ASV")

saveRDS(merged_physeq, file = "non_normalised_physeq_file.rds")
```

# 22. CSS normalization

Rarefaction was not performed due to substantial variation in sampling depth across different timepoints, which would result in significant read loss. Sequencing counts were therefore normalized using Cumulative Sum Scaling (CSS).
```{r, eval = FALSE}

normalised_physeq <- phyloseq_transform_css(merged_physeq)
```

# 23. Saving phyloseq object

This phyloseq object will be used in further statistical analysis, and visualization
```{r, eval = FALSE}

saveRDS(normalised_physeq, file = "normalised_physeq_file.rds")
```

# 24. Session information
```{r}

sessionInfo()
```

