---
title: "Reproducible downstream microbiome analysis"
author: "Stefanie Katharina Thaqi and Srotoswini Joardar"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Scope

This workflow reproduces the downstream microbiome analyses reported in the manuscript.
It starts from a processed, CSS-normalized `phyloseq` object containing the filtered ASV table, sample metadata, SILVA taxonomy, and a phylogenetic tree. Sequence processing, contaminant removal, filtering, tree construction, and CSS normalization were performed as described in the manuscript and is available in a separate rmd file (Amplicon pre processing.rmd) available in GitHub.

The analyses include alpha diversity, Bray-Curtis NMDS and PERMANOVA, genus-level community composition, and abundance weighted beta nearest taxon index (βNTI).

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(phyloseq)
library(microbiome)
library(vegan)
library(ape)
library(picante)
library(tidyverse)
library(patchwork)
library(emmeans)
library(rstatix)
library(ggpubr)
library(reshape2)
library(scales)

set.seed(42)


treatment_levels <- c("Control", "Autoclaved", "Bacillus", "Pseudomonas", "Mix")
time_levels <- c("T1", "T2", "T3")
selected_timepoints <- c("T2", "T3")

# Colours used in the submitted figures.
treatment_colors <- c(
  "Control" = "#9DA9A2",
  "Autoclaved" = "#E0CA72",
  "Bacillus" = "#DF857F",
  "Pseudomonas" = "#52AFE0",
  "Mix" = "#A868E8"
)

time_shapes <- c("T1" = 21, "T2" = 22, "T3" = 24)

base_theme <- theme_classic(base_size = 14) +
  theme(
    text = element_text(family = "Arial", color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```

# Load the processed phyloseq object

```{r load-data}
# The repository must contain data/root_ps.RData with an object named root_ps.
load("data/root_ps.RData")
ps <- root_ps
rm(root_ps)

metadata <- as.data.frame(sample_data(ps))

stopifnot(
  all(metadata$Treatment %in% treatment_levels),
  all(metadata$Time_point %in% time_levels)
)

metadata$Treatment <- factor(metadata$Treatment, levels = treatment_levels)
metadata$Time_point <- factor(metadata$Time_point, levels = time_levels)

sample_data(ps) <- sample_data(metadata)
metadata_df <- metadata %>% rownames_to_column("SampleID")

ps
```

# Alpha diversity

Observed richness was square-root transformed. Treatment, sampling time point, and their interaction were tested using linear models. Tukey-adjusted treatment contrasts were calculated within each sampling time point.

```{r alpha-diversity}
observed_df <- estimate_richness(ps, measures = "Observed") %>%
  as.data.frame() %>%
  rownames_to_column("SampleID")

pielou_df <- microbiome::alpha(ps, index = "evenness_pielou") %>%
  as.data.frame()
colnames(pielou_df) <- "Pielou"
pielou_df <- pielou_df %>% rownames_to_column("SampleID")

alphadiv_df <- observed_df %>%
  left_join(pielou_df, by = "SampleID") %>%
  left_join(metadata_df, by = "SampleID")

model_observed <- lm(sqrt(Observed) ~ Treatment * Time_point, data = alphadiv_df)
model_pielou <- lm(Pielou ~ Treatment * Time_point, data = alphadiv_df)

anova_observed <- anova(model_observed)
anova_pielou <- anova(model_pielou)

observed_contrasts <- emmeans(
  model_observed, pairwise ~ Treatment | Time_point, adjust = "tukey"
)$contrasts %>% summary() %>% as.data.frame()

pielou_contrasts <- emmeans(
  model_pielou, pairwise ~ Treatment | Time_point, adjust = "tukey"
)$contrasts %>% summary() %>% as.data.frame()

anova_observed
observed_contrasts
anova_pielou
pielou_contrasts
```

```{r alpha-plots, fig.width=9, fig.height=7}
observed_plot <- ggplot(alphadiv_df, aes(Treatment, Observed, fill = Treatment)) +
  geom_boxplot(width = 0.7, outlier.shape = 21, outlier.fill = "black") +
  facet_wrap(~Time_point) +
  scale_fill_manual(values = treatment_colors, drop = FALSE) +
  labs(x = NULL, y = "Observed number of ASVs") +
  base_theme + theme(legend.position = "none")

pielou_plot <- ggplot(alphadiv_df, aes(Treatment, Pielou, fill = Treatment)) +
  geom_boxplot(width = 0.7, outlier.shape = 21, outlier.fill = "black") +
  facet_wrap(~Time_point) +
  scale_fill_manual(values = treatment_colors, drop = FALSE) +
  labs(x = NULL, y = "Pielou evenness") +
  base_theme + theme(legend.position = "none")

alpha_plot <- observed_plot / pielou_plot
alpha_plot
```

# Bray-Curtis beta diversity

The input ASV table is already CSS-normalized. No additional normalization is applied before Bray-Curtis dissimilarity. The submitted analysis compares T2 and T3. Sequential PERMANOVA terms are tested in the order Treatment, Time point, and their interaction to reproduce the submitted analysis.

```{r beta-diversity, fig.width=7, fig.height=5}
ps_beta <- subset_samples(ps, Time_point %in% selected_timepoints)
ps_beta <- prune_samples(sample_sums(ps_beta) > 0, ps_beta)
ps_beta <- prune_taxa(taxa_sums(ps_beta) > 0, ps_beta)

metadata_beta <- as.data.frame(sample_data(ps_beta))
metadata_beta$Treatment <- factor(metadata_beta$Treatment, levels = treatment_levels)
metadata_beta$Time_point <- factor(metadata_beta$Time_point, levels = selected_timepoints)
metadata_beta_df <- metadata_beta %>% rownames_to_column("SampleID")

set.seed(42)
nmds_t2_t3 <- ordinate(
  ps_beta, method = "NMDS", distance = "bray",
  trymax = 200, autotransform = FALSE, trace = FALSE
)

nmds_t2_t3$stress

nmds_t2_t3_df <- vegan::scores(nmds_t2_t3, display = "sites") %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  left_join(metadata_beta_df, by = "SampleID")

nmds_t2_t3_plot <- ggplot(
  nmds_t2_t3_df,
  aes(NMDS1, NMDS2, fill = Treatment, shape = Time_point)
) +
  geom_point(size = 4, color = "black", stroke = 0.6, alpha = 0.9) +
  stat_ellipse(
    aes(group = Time_point), color = "black", linetype = 2,
    linewidth = 0.6, level = 0.95, type = "t"
  ) +
  scale_fill_manual(values = treatment_colors) +
  scale_shape_manual(values = time_shapes[selected_timepoints]) +
  labs(x = "NMDS1", y = "NMDS2", fill = "Treatment", shape = "Time point") +
  base_theme

nmds_t2_t3_plot
```

```{r permanova}
bray_t2_t3 <- phyloseq::distance(ps_beta, method = "bray")
metadata_t2_t3 <- metadata_beta[labels(bray_t2_t3), , drop = FALSE]

set.seed(42)
permanova_t2_t3 <- adonis2(
  bray_t2_t3 ~ Treatment * Time_point,
  data = metadata_t2_t3,
  permutations = 999,
  by = "terms"
)
permanova_t2_t3

dispersion_group <- interaction(
  metadata_t2_t3$Treatment, metadata_t2_t3$Time_point, drop = TRUE
)
dispersion_t2_t3 <- betadisper(bray_t2_t3, group = dispersion_group)
set.seed(42)
dispersion_test_t2_t3 <- permutest(dispersion_t2_t3, permutations = 999)
dispersion_test_t2_t3
```

# Genus-level community composition

For visualization only, CSS-normalized abundances are converted to within-sample relative abundances. The 15 most abundant genera are selected separately for each sampling time point; all remaining genera are grouped as `Other`.

```{r community-composition, fig.width=11, fig.height=5}
ps_composition <- transform_sample_counts(ps, function(x) x / sum(x))

tax_matrix <- as(tax_table(ps_composition), "matrix")
missing_genus <- is.na(tax_matrix[, "Genus"]) | trimws(tax_matrix[, "Genus"]) == ""
tax_matrix[missing_genus, "Genus"] <- "Unclassified"
tax_table(ps_composition) <- tax_table(tax_matrix)

ps_genus <- tax_glom(ps_composition, taxrank = "Genus", NArm = FALSE)

genus_sample_df <- psmelt(ps_genus) %>%
  transmute(
    Sample,
    Treatment = factor(Treatment, levels = treatment_levels),
    Time_point = factor(Time_point, levels = time_levels),
    Genus = as.character(Genus),
    Abundance
  ) %>%
  group_by(Sample, Treatment, Time_point, Genus) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

top_genera_by_time <- genus_sample_df %>%
  group_by(Time_point, Genus) %>%
  summarise(MeanAbundance = mean(Abundance), .groups = "drop") %>%
  group_by(Time_point) %>%
  slice_max(MeanAbundance, n = 15, with_ties = FALSE) %>%
  ungroup() %>%
  select(Time_point, Genus) %>%
  mutate(is_top = TRUE)

genus_plot_df <- genus_sample_df %>%
  left_join(top_genera_by_time, by = c("Time_point", "Genus")) %>%
  mutate(Genus = if_else(is_top %in% TRUE, Genus, "Other")) %>%
  group_by(Sample, Treatment, Time_point, Genus) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Treatment, Time_point, Genus) %>%
  summarise(MeanAbundance = mean(Abundance), .groups = "drop")

genus_order <- genus_plot_df %>%
  filter(Genus != "Other") %>%
  group_by(Genus) %>%
  summarise(OverallMean = mean(MeanAbundance), .groups = "drop") %>%
  arrange(desc(OverallMean)) %>%
  pull(Genus)

genus_levels <- c(genus_order, "Other")
genus_plot_df$Genus <- factor(genus_plot_df$Genus, levels = genus_levels)

genus_palette <- setNames(
  grDevices::hcl.colors(length(genus_levels), palette = "Dynamic"),
  genus_levels
)
genus_palette["Other"] <- "grey60"

community_plot <- ggplot(
  genus_plot_df,
  aes(Treatment, MeanAbundance, fill = Genus)
) +
  geom_col(width = 0.8, color = "black", linewidth = 0.2) +
  facet_wrap(~Time_point, nrow = 1) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(0, 1), expand = expansion(mult = c(0, 0))
  ) +
  scale_x_discrete(drop = FALSE) +
  scale_fill_manual(values = genus_palette, drop = FALSE) +
  labs(x = NULL, y = "Mean relative abundance (%)", fill = "Genus") +
  base_theme +
  theme(legend.text = element_text(size = 10))

community_plot
```

# Phylogenetic turnover (βNTI)

βNTI is calculated separately for each Treatment × Time point subset. Positive values above +2 indicate variable selection, values below -2 indicate homogeneous selection, and values between -2 and +2 fall within the null expectation. Pairwise Wilcoxon tests reproduce the analysis used for the submitted figure. Because pairwise βNTI values can share samples, inferential results should be interpreted cautiously.

```{r bnti-functions}
prune_to_tree_and_data <- function(ps_object) {
  ps_object <- prune_samples(sample_sums(ps_object) > 0, ps_object)
  ps_object <- prune_taxa(taxa_sums(ps_object) > 0, ps_object)
  taxa_keep <- intersect(taxa_names(ps_object), phy_tree(ps_object)$tip.label)
  ps_object <- prune_taxa(taxa_keep, ps_object)

  if (nsamples(ps_object) < 2) stop("At least two samples are required for βNTI.")
  if (ntaxa(ps_object) < 2) stop("At least two tree-matched taxa are required for βNTI.")
  ps_object
}

calculate_bnti <- function(ps_sub, n_reps = 999) {
  ps_sub <- prune_to_tree_and_data(ps_sub)
  otu <- as(otu_table(ps_sub), "matrix")
  if (taxa_are_rows(ps_sub)) otu <- t(otu)

  tree <- phy_tree(ps_sub)
  cophe <- cophenetic(tree)
  otu <- otu[, colnames(otu) %in% rownames(cophe), drop = FALSE]
  otu <- otu[rowSums(otu) > 0, colSums(otu) > 0, drop = FALSE]
  cophe <- cophe[colnames(otu), colnames(otu), drop = FALSE]

  observed <- as.matrix(comdistnt(otu, cophe, abundance.weighted = TRUE))

  null_array <- replicate(n_reps, {
    shuffled_tree <- tree
    shuffled_tree$tip.label <- sample(tree$tip.label)
    shuffled_cophe <- cophenetic(shuffled_tree)
    shuffled_cophe <- shuffled_cophe[colnames(otu), colnames(otu), drop = FALSE]
    as.matrix(comdistnt(otu, shuffled_cophe, abundance.weighted = TRUE))
  }, simplify = "array")

  null_mean <- apply(null_array, c(1, 2), mean, na.rm = TRUE)
  null_sd <- apply(null_array, c(1, 2), sd, na.rm = TRUE)
  null_sd[null_sd == 0] <- NA_real_

  bnti <- (observed - null_mean) / null_sd
  diag(bnti) <- NA_real_
  bnti[!is.finite(bnti)] <- NA_real_
  bnti
}
```

```{r bnti-calculation, message=TRUE}
set.seed(42)
bnti_list <- list()

for (tp in time_levels) {
  for (trt in treatment_levels) {
    ps_subset <- subset_samples(ps, Time_point == tp & Treatment == trt)
    ps_subset <- prune_samples(sample_sums(ps_subset) > 0, ps_subset)
    ps_subset <- prune_taxa(taxa_sums(ps_subset) > 0, ps_subset)

    if (nsamples(ps_subset) < 2 || ntaxa(ps_subset) < 2) {
      message("Skipped: ", tp, " | ", trt)
      next
    }

    message("Calculating βNTI: ", tp, " | ", trt)
    bnti_list[[paste(tp, trt, sep = "|")]] <- calculate_bnti(ps_subset, n_reps = 999)
  }
}

bnti_df <- bind_rows(lapply(names(bnti_list), function(key) {
  key_parts <- strsplit(key, "\\|")[[1]]
  reshape2::melt(
    bnti_list[[key]],
    varnames = c("Sample1", "Sample2"),
    value.name = "bNTI"
  ) %>%
    mutate(Sample1 = as.character(Sample1), Sample2 = as.character(Sample2)) %>%
    filter(Sample1 < Sample2) %>%
    mutate(Time_point = key_parts[1], Treatment = key_parts[2])
})) %>%
  mutate(
    Treatment = factor(Treatment, levels = treatment_levels),
    Time_point = factor(Time_point, levels = time_levels),
    Assembly_category = case_when(
      bNTI > 2 ~ "Variable selection",
      bNTI < -2 ~ "Homogeneous selection",
      TRUE ~ "Within null expectation"
    )
  )

bnti_summary <- bnti_df %>%
  group_by(Time_point, Treatment) %>%
  summarise(
    n_pairwise_comparisons = sum(!is.na(bNTI)),
    median_bNTI = median(bNTI, na.rm = TRUE),
    mean_bNTI = mean(bNTI, na.rm = TRUE),
    sd_bNTI = sd(bNTI, na.rm = TRUE),
    proportion_variable_selection = mean(bNTI > 2, na.rm = TRUE),
    proportion_homogeneous_selection = mean(bNTI < -2, na.rm = TRUE),
    proportion_within_null_expectation = mean(abs(bNTI) <= 2, na.rm = TRUE),
    .groups = "drop"
  )

bnti_category_summary <- bnti_df %>%
  filter(!is.na(bNTI)) %>%
  count(Time_point, Treatment, Assembly_category, name = "n") %>%
  group_by(Time_point, Treatment) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

bnti_summary
```

```{r bnti-tests}
bnti_comparisons <- list(
  c("Control", "Pseudomonas"),
  c("Control", "Mix"),
  c("Autoclaved", "Pseudomonas"),
  c("Bacillus", "Pseudomonas")
)

bnti_wilcoxon <- bnti_df %>%
  filter(!is.na(bNTI)) %>%
  group_by(Time_point) %>%
  pairwise_wilcox_test(
    bNTI ~ Treatment,
    comparisons = bnti_comparisons,
    p.adjust.method = "BH"
  )

bnti_wilcoxon_plot <- bnti_wilcoxon %>%
  filter(p.adj <= 0.05) %>%
  add_xy_position(x = "Treatment", dodge = 0)
```

```{r bnti-plots, fig.width=9, fig.height=4.5}
bnti_plot <- ggplot(bnti_df, aes(Treatment, bNTI, fill = Treatment)) +
  geom_boxplot(width = 0.7, outlier.shape = 21, outlier.fill = "black") +
  geom_jitter(width = 0.2, size = 2, alpha = 0.4, color = "black") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", color = "red") +
  scale_fill_manual(values = treatment_colors) +
  facet_wrap(~Time_point, nrow = 1) +
  labs(x = NULL, y = expression(beta * "NTI")) +
  base_theme + theme(legend.position = "none")

if (nrow(bnti_wilcoxon_plot) > 0) {
  bnti_plot <- bnti_plot + stat_pvalue_manual(
    bnti_wilcoxon_plot,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.6
  )
}

assembly_plot <- ggplot(
  bnti_category_summary,
  aes(Treatment, proportion, fill = Assembly_category)
) +
  geom_col(color = "black", linewidth = 0.2) +
  facet_wrap(~Time_point, nrow = 1) +
  labs(
    x = NULL,
    y = "Proportion of pairwise comparisons",
    fill = "Assembly category"
  ) +
  base_theme

bnti_plot
assembly_plot
```

# Session information

```{r session-info}
sessionInfo()
```
